Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of neutron star - black hole binary observations

### 2017-01-13 Paper/journal_submission/paper.tex
incorporating MP's response to responses to PRD

We start with calculating $\arr_{\mchirp}$ as a function of various source
parameters and show it in Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}.
$\mchirp$ is the leading order
mass combination that affects the GW strain emitted by compact binaries as they
spiral in, and is therefore determined the most precisely. We notice
immediately that for $\rho\leq 30$ the systematics are well under control
and we can obtain reliable chirp mass estimates for NSBH signals using BBH
templates.
% 
For louder and less likely SNRs ($\rho\simeq 50$), we find that
$\arr_{\mchirp}$ can become comparable to unity, but only if the BH has
prograde spin $\chibh\gtrsim 0.4$, {\it and} the true NS tidal deformability
is large enough, s.t. $\lambdans \gtrsim 1000$.
% 
We therefore conclude that only for very loud signals, with
$\rho\gtrsim 50-70$, will the inclusion of tidal terms in template
models improve $\mchirp$ estimation. For lower SNRs, inclusion of new
physical content in templates will instead get washed out by detector noise.
% 
In addition, we also note that $\arr_{\mchirp}\geq 0$ always,
i.e. $\mchirp$ is always being over-estimated. This is to be expected since
the tidal deformation of the NS drains energy from the orbit faster than in
the BBH case, and the disruption of NS close to merger reduces GW signal
power at high frequencies. Both of these effects make the resulting signal
resemble a BBH signal of higher chirp mass.

@@

$\mbh\geq 5M_\odot$ (right panel).

@@

{\it Do all events matter?} In Ref.~\cite{Lackey2014} the authors demonstrate
that the overwhelming majority of information about the NS equation of state
comes from the loudest $\sim 5$ events in the case of binary neutron star
detections, and not from the combined effect of a large number of low-SNR
systems.
The question naturally arises if the same is true for NSBH sources as well.
Therefore, in Fig.~\ref{fig:TT_LambdaError_vs_N_L500_2000_CI90_0_AllInOne_SNRSorted}
we re-evaluate the accumulation of information with each successive NSBH 
detection, sorting the events in each population by their SNR instead of 
simulated chronology. We find the same qualitative behavior as in the case of 
BNSs~\cite{Lackey2014}. 
Whether or not there is an astrophysical mass gap, we find
that the bulk of tidal information will be furnished by the loudest $5-10$
NSBH detections of aLIGO detectors. This result is especially encouraging
to NR follow-up efforts for GW detections, as we now know that only a handful
of loudest NSBH events will merit full numerical-GR follow-up simulations.

@@

We probe two astrophysical paradigms, one that allows for BH masses to lie
within the astrophysical mass-gap (paradigm A), and one that does not (paradigm
B).
% 
{\it For paradigm A}, we find the following: (i) for the softer equations of
state that result in less deformable neutron stars, $15-20$ detections bring
the measured probability distribution for $\lambdans$ entirely within the prior,
which ensures that the median $\lambdans$ tracks the true value to within $10\%$.
(ii) For NSBH populations with more deformable NSs ($\lambdans> 1000$),
the same is achievable within as few as $10$ (or $15$ at most) realistic
observations. (iii) The statistical uncertainty associated with $\lambdans$
measurement can be restricted to be within $\pm50\%$ using $10-20$ observations
when $\lambdans> 1000$), and using $25-40$ observations for softer equations
of state. All of the above is possible within a few years of design
aLIGO operation~\cite{Abadie:2010cfa}, if astrophysical BHs are allowed
masses $< 5M_\odot$ (i.e. in the mass-gap). However, further
restricting $\lambdans$ will require $50+$ NSBH observations.
% 
{\it For paradigm B}, we find the information accumulation to be somewhat slower.
While the quantitative inferences for populations with $\lambdans>1000$ are
not affected significantly, we find that $\lambdans< 1000$ populations require 
$10-20\%$ more events to attain the same measurement accuracy as under
paradigm A. In either case, the accumulation of information from the general
NSBH population will likely be slower than from BNS inspirals~\cite{Mandel:2009pc,
Lackey2014,Wade:2014vqa,Agathos:2015a}, depending on the mass distribution of 
stellar-mass black holes. Though, template models for the latter may be more
uncertain due to missing point-particle PN terms at orders comparable to
tidal terms~\cite{Lackey2014}.
% 
We conclude that within as few as $20-30$ observations of disruptive NSBH
mergers, aLIGO will begin to place interesting bounds on NS deformability.
This, amongst other things, will allow us to rank different equations of 
state for neutron star matter from most to least likely, within a few years'
detector operation.
% 
We also find that, within this population, the loudest $\sim 5-10$ events
will provide us with most of the tidal information, and will therefore
merit full NR follow-up.
% 
Our methods and results are detailed in Sec.~\ref{s1:multiple_observations}.

>>>>

We start with calculating $\arr_{\mchirp}$ as a function of various source
parameters and show it in Fig.~\ref{fig:TN_chirpMassBias_vs_Lambda_SNR}.
$\mchirp$ is the leading order
mass combination that affects the GW strain emitted by compact binaries as they
spiral in, and is therefore determined the most precisely. We notice
immediately that for $\rho\leq 30$ the systematics are well under control
and we can obtain reliable chirp mass estimates for NSBH signals using BBH
templates.
% 
For louder and less likely SNRs ($\rho\simeq 50$), we find that
$\arr_{\mchirp}$ can become comparable to unity, but only if the BH has
prograde spin $\chibh\gtrsim 0.4$, {\it and} the true NS tidal deformability
is large enough, s.t. $\lambdans \gtrsim 1000$.
% 
We therefore conclude that only for very loud signals, with
$\rho\gtrsim 50-70$, will the inclusion of tidal terms in template
models improve $\mchirp$ estimation. For lower SNRs, inclusion of new
physical content in templates will instead get washed out by detector noise.
% 
In addition, we also note that $\arr_{\mchirp}\geq 0$ always,
i.e. $\mchirp$ is always being over-estimated. This is to be expected since
the tidal deformation of the NS drains energy faster from the orbit during
inspiral (as compared to the BBH case), and its disruption close to merger
reduces GW signal power at high frequencies. Both of these effects make the
resulting signal resemble a BBH signal of higher chirp (or total) mass, although
we expect the latter effect to be dominant~\cite{Pannarale:2011pk}.

@@

$\mbh\geq 5M_\odot$ (right panel). Left inset zooms
in on the main figure for the first $15$ events. Right inset
shows the actual (ensemble mean) SNR value for each event. We find that
events with $\rho\gtrsim 20-30$ provide the bulk of tidal information
in our analysis.

@@

{\it Do all events matter?} In Ref.~\cite{Lackey2014} the authors demonstrate
that the overwhelming majority of information about the NS equation of state
comes from the loudest $\sim 5$ events in the case of binary neutron star
detections, and not from the combined effect of a large number of low-SNR
systems.
The question naturally arises if the same is true for NSBH sources as well.
Therefore, in Fig.~\ref{fig:TT_LambdaError_vs_N_L500_2000_CI90_0_AllInOne_SNRSorted}
we re-evaluate the accumulation of information with each successive NSBH 
detection, sorting the events in each population by their SNR instead of 
simulated chronology. We find the same qualitative behavior as in the case of 
BNSs~\cite{Lackey2014}. 
Whether or not there is an astrophysical mass gap, we find
that the bulk of tidal information will be furnished by the loudest $5-10$
NSBH detections of aLIGO detectors. {\it This result is especially encouraging
to NR follow-up efforts for GW detections, as we now know that only a handful
of loudest NSBH events (with SNRs $\rho\gtrsim 20-30$) are the ones that may merit
full numerical-GR + magnetohydrodynamical follow-up simulations.}

@@

We probe two astrophysical paradigms, one that allows for BH masses to lie
within the astrophysical mass-gap (paradigm A), and one that does not (paradigm
B).
% 
{\it For paradigm A}, we find the following: (i) for the softer equations of
state that result in less deformable neutron stars, $15-20$ detections bring
the measured probability distribution for $\lambdans$ entirely within the prior,
which ensures that the median $\lambdans$ tracks the true value to within $10\%$.
(ii) For NSBH populations with more deformable NSs ($\lambdans> 1000$),
the same is achievable within as few as $10$ (or $15$ at most) realistic
observations. (iii) The statistical uncertainty associated with $\lambdans$
measurement can be restricted to be within $\pm50\%$ using $10-20$ observations
when $\lambdans> 1000$), and using $25-40$ observations for softer equations
of state. All of the above is possible within a few years of design
aLIGO operation~\cite{Abadie:2010cfa}, if astrophysical BHs are allowed
masses $< 5M_\odot$ (i.e. in the mass-gap). However, further
restricting $\lambdans$ will require $50+$ NSBH observations.
% 
{\it For paradigm B}, we find the information accumulation to be somewhat slower.
While the quantitative inferences for populations with $\lambdans>1000$ are
not affected significantly, we find that $\lambdans< 1000$ populations require 
$10-20\%$ more events to attain the same measurement accuracy as under
paradigm A. In either case, the accumulation of information from the general
NSBH population will likely be slower than from BNS inspirals~\cite{Mandel:2009pc,
Lackey2014,Wade:2014vqa,Agathos:2015a}, depending on the mass distribution of 
stellar-mass black holes. Though, template models for the latter may be more
uncertain due to missing point-particle PN terms at orders comparable to
tidal terms~\cite{Lackey2014}.
% 
We conclude that within as few as $20-30$ observations of disruptive NSBH
mergers, aLIGO will begin to place interesting bounds on NS deformability.
This, amongst other things, will allow us to rank different equations of 
state for neutron star matter from most to least likely, within a few years'
detector operation.
% 
We also find that, within this population, the loudest $5-10$ events (with SNRs
$\rho\gtrsim 20-30$) will provide us with most of the tidal information,
and will therefore merit full NR follow-up.
% 
Our methods and results are detailed in Sec.~\ref{s1:multiple_observations}.
